Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2021-09-16 paper/latex/NetworkNeuroscience/rmt.tex
add cover letter

\documentclass[NETN,manuscript]{stjour-new}
\articletype{Research}
\usepackage{graphicx}
\graphicspath{ {figs/} }
% \input{preamble.tex}
% \documentclass{article}
% \usepackage[utf8]{inputenc}
% \usepackage{geometry}
% \usepackage{amsmath}
% \usepackage{graphicx}
% \usepackage[style=apa]{biblatex}
% \usepackage{natbib}
% \usepackage[font=footnotesize,labelfont=bf]{caption}
% \addbibresource{rmt.bib}
% \bibliographystyle{plain}
% \bibliography{rmt}
\begin{document}

@@

\date{\today}

% \maketitle

% \keywords{random matrix theory, feature extraction, functional connectivity,
% functional magnetic resonance imaging, voxelwise connectivity, predictive analysis}

\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse

    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make code publicly available to extract
    these features from a wide variety of data. We find preliminary evidence suggesting that
    RMT-inspired features may have unique potential in analyses of fMRI functional connectivity.
\end{abstract}

@@

In functional magnetic resonance imaging (fMRI), changes in the blood-oxygenation-level-dependent
(BOLD) signals are related to neural activity. It is common to investigate statistical relationships
between the BOLD signals through functional connectivity analyses, where correlations between
collections of these signals are examined to infer connections between different voxels or regions
of interest (ROIs) within the brain.

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
\citep{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}.
This complexity suggests fMRI is a candidate to be studied using Random Matrix Theory (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics \citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has been applied to atmospheric
physics \citep{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations \cite{plerouRandomMatrixApproach2002} , social networks \cite{jalanUncoveringRandomnessSuccess2014}, random networks \cite{bandyopadhyayUniversalityComplexNetworks2007},
network-formation in liquids \citep{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}  and amorphous clusters \citep{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}. Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships \citep{bhadolaTargetingFunctionalMotifs2016},
synchrony in epileptic seizures \citep{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein interactions both in different species \citep{agrawalQuantifyingRandomnessProtein2014} and breast cancer \citep{raiRandomnessPreservedPatterns2015}.
\end{sloppypar}

@@

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states \citep{wangSpectralPropertiesTemporal2015}, between
subjects with and without ADHD \citep{wangRandomMatrixTheory2016}, and between pain and non-pain
states \citep{matharooSpontaneousBackpainAlters2020}. Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE.

These findings suggest that certain aspects of psychological processes might be characterized, in
part, by features computed from the eigenvalues of fMRI correlation matrices, and that these
features might vary in an interpretable manner across psychological processes. If this is the case,
RMT could have unique potential in characterizing the functioning of the human brain.

Here, we use a novel, voxel-based approach, and expand the applications of RMT to analyze fMRI scans
from  diverse datasets1. We extract RMT-inspired whole-brain features from voxelwise functional
connectivity data, and assess the predictive value of these features. We also examine in detail the
robustness of the RMT methodology to various common analytic choices.

@@

As our primary intention is to examine the potential utility of RMT-derived features across a wide
variety of data, and not to decisively put forward or refute any specific hypotheses relating RMT to
psychological processes, we examine a wide selection of fMRI datasets available on the OpenNeuro
platform [36].  This allows for a broad assessment of the potential of voxelwise RMT features.

However, since previous studies employing RMT for feature extraction
\citep{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020} have partially interpreted
their findings with respect to attentional processes, we have elected to use datasets that involve
attention in some way. This includes measures of attention or vigilance, or comparisons between a
condition involving more attention and focus (e.g. a task state) or conditions which should involve
considerably less attention or focus (e.g. resting-state).

@@

\paragraph{Trait Attention}
The PANAS-X \citep{watsonPANASXManualPositive1994} is a 9-point Likert-type self-report measure
that queries the extent to which subjects agree particular affective terms have applied to them over
a recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) sum-score was constructed from
the scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects with a z-score greater / less than zero on this
sum-score were placed in the high / low ``trait attention'' category

\paragraph{Task Attention}
The Conjunctive Continuous Performance Task (CCPT;
\cite{shalevConjunctiveContinuousPerformance2011} is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\paragraph{Vigilance}
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \cite{gorgolewskiCorrespondenceIndividualDifferences2014}. One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in Fig. and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{table:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

@@

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast \citep{kunduMultiechoFMRIReview2017}. In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's ith signal value
from echo j as , then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in Fig. and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{table:1}. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
\citep[ANT;][]{fanActivationAttentionalNetworks2005}. Subgroup details are available in Table \ref{table:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in Fig. and tables with the label ``LEARN'') is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components \citep[see figure 1
of][]{schapiroHumanHippocampalReplay2018}. After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of
the analysis of this paper. Resting state scans were included in the low-attention ``rest''
subgroup.

@@

The RMT features extracted here are based on functional correlations rather than ROI activations,
and are completely insensitive to voxel location (spatial re-arrangement of an image’s voxels does
not change the eigenvalues of the correlation matrix). Thus RMT features summarize all voxel-wise
correlations in a whole-brain and location-free manner, which brings into question some of the usual
motivations for performing spatial smoothing, or registration to identify ROIs. In addition, if
image dimensions are identical between subjects, the RMT features (by virtue of their definition and
the unfolding procedure) already allow for direct comparisons. Thus, the usual motivations for
registration are also no longer clear when extracting RMT features.

@@

where \(\sigma^2\) is the true variance of the distribution of each i.i.d. entry (equal to 1 in the
case of correlation matrices), and where \(r = N / T\), for matrices of dimension \(N \times T\).
While real-world data will not, in general, be i.i.d., the extent to which spectra overlap with or
are well-described by the MP distribution has been used to guide analytical decisions regarding the
relevance of empirically observed eigenvalues. That is, since the majority of the eigenvalues of
random correlation matrices will fall within \([\lambda_{-}, \lambda_{+}]\), eigenvalues observed to
fall within this region are typically considered to be the result of noise
\citep{veraartDenoisingDiffusionMRI2016,veraartDiffusionMRINoise2016,ulfarssonDimensionEstimationNoisy2008,macmahonCommunityDetectionCorrelation2015}.
We examine as a potential feature the proportion of eigenvalues that fall within this range, and
call this the noise ratio. In addition, we calculate a broader, shifted noise ratio, based on the
adjustments of \cite{almogUncoveringFunctionalSignature2019} which takes into account the larger
degree of correlation in empirical data.

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table \ref{table:1}. The various RMT
features mentioned above were then extracted via the empyricalRMT library [64] developed and made
publicly available as part of this study, as well as via specific code for this paper (which is
documented and available online \citep{bergerderekm.DMBergerRandommatrixfmriV02020}.

@@

A wide variety of standard, out-of-the-box classifiers available in the Python scikit-learn library
\citep{pedregosaScikitlearnMachineLearning2011} were then employed to gauge the predictive value of
the RMT features. To allow for a wide range of flexibilities, classifiers employing k-nearest
neighbours (KNN), with k values in \(\{1, 2, 3, \dots, 10, 20\}\), logistic regression (LR), linear
discriminant analysis (LDA), and a support-vector machine (SVM) were employed. Default parameter
values were used when possible (see \cite{bergerderekm.DMBergerRandommatrixfmriV02020} for precise
function calls used).

@@

\subsection{Overview}
Due to space limitations, we highlight only key results here. Extensive overviews of the full
results are available in the supplementary materials. Visual summaries of the prediction accuracies
for all features and analytic choices are presented in Fig. S1--4. Prediction accuracies for only the
multivariate RMT features are summarized in Fig. S5--7, and the distributions of the scalar features
are presented in Fig. S8--10. Prediction accuracies for the scalar features across analytic choices
are summarized in Fig. S11--12.

@@

With the exception of the LEARNING dataset, the RMT-inspired features all demonstrated some
predictive utility (Fig. S1-4). For example, there were strong differences in RMT features between
task and rest conditions for the REFLECT dataset, but the features were of variable predictive
utility in the PSYCH dataset, sometimes outperforming the naïve classifier, and sometimes not.
Although the RMT features were useful in the comparison involving duloxetine and controls (no pain)
in the OSTEO dataset, these features mostly did not aid in classifying between any of the other
possible pairings of subgroups in this dataset (Fig. S1-4).

Although the performance of classifiers trained on RMT features was variable, there was a tendency
for the raw and largest eigenvalues to be the most predictively useful features, especially in the
REFLECT and PARK data in discriminating between Parkinson's and controls (Fig. S5-7. S13; but see
Discussion).

\subsection{Raw Features}
As evident from Fig. S8-10, the scalar raw features (largest eigenvalue, noise ratio, and shifted
noise ratio) varied more within than between subgroups. They were not generally useful for
classifying conditions (Tables S15-17), with two or three notable exceptions.

The noise ratio (but not shifted noise ratio) differed consistently between the duloxetine and
control groups in the OSTEO data, allowing for prediction accuracies of up to 72\% compared to 51\%
for the naive classifier (Fig. S11-12 top-right panel, Table S5). The largest eigenvalue was also
useful in distinguishing between subjects self-reporting high versus low vigilance across sessions:
accuracies up to 64\% and 65\% (compared to naïve accuracies of 55\% and 50\%) for the first and
second sessions, respectively (Table S6). Finally, for the REFLECT data, task and rest were
perfectly distinguishable with various features and with some classifiers.

The sorted, raw eigenvalues, however, more consistently contained information that allowed some
separation of subgroups. Averaging across all datasets and comparisons, the raw eigenvalues
consistently demonstrated predictive utility (Tables S7--8). As visible in Fig. S13, the eigenvalues
had the most consistent predictive value for distinguishing individuals with Parkinson's from
controls, or distinguishing those with osteopathic pain taking duloxetine from controls (``nopain'' in
Fig. S13). The raw eigenvalues sometimes had predictive value for the PSYCH data, most notably in
comparing those differing in self-reported vigilance and high vs. low scorers in the attention task
(Table S7--8).

@@

\paragraph{Role of Classifer}
The raw eigenvalues generally had the most predictive utility when employing either an SVM
classifier or KNN classifier with \(k > 3\) (Table S11, first row). When the raw eigenvalues were
first min-max normalized prior to prediction, an SVM classifier was clearly the most consistent in
predicting subgroup memberships (Table S12), however without normalization, KNN with \(k = 8\) or
\(k = 9\) generally best predicted subgroup membership across datasets (Table S11).

@@

The predictive utility of the unfolded RMT features varied considerably across datasets and
classifiers. However, some results were robust: for example, the spectral rigidity and level number
variance strongly differentiated controls from individuals prescribed duloxetine for osteopathic
pain, and this differentiation persisted regardless of unfolding degree or trimming (Fig. 5-8, top-right panels).

@@

For other datasets, the predictive utility of the unfolded features was highly dependent on the
unfolding degree, and sometimes trim proportions. These differences are illustrated in detail for
the Parkinson's data in Fig. S14. For example, with minimal trimming, subgroup overlap in spectral
rigidity was minimized for unfolding degrees 5 and 9 (Fig. S14, first row), but with greater
trimming, this was not the case (Fig. S14, second row). For the level number variance with minimal
trimming, subgroup overlap appeared smaller for lower unfolding degrees (Fig. S14, third row),
however, with a greater degree of trimming, separation was clearer for slightly larger degrees (Fig.
S14, fourth row).

@@

\paragraph{Role of Classifier}. Across all datasets, unfoldings, and trimmings, the spectral rigidity was
generally best able to separate between subgroups when either k was large (10 or 20) in KNN, or
Logistic Regression or SVM was employed (Table S13). The pattern was similar for the level number
variance, although the predictive utility of the level number variance was, overall, slightly more
variable than the spectral rigidity (Table S4).

@@

\section{Conclusion}

>>>>

\documentclass[NETN,manuscript]{stjour-new}
\articletype{Research}
\usepackage{graphicx}
\usepackage{cleveref}
\graphicspath{ {figs/} }
% \input{preamble.tex}
% \documentclass{article}
% \usepackage[utf8]{inputenc}
% \usepackage{geometry}
% \usepackage{amsmath}
% \usepackage{graphicx}
% \usepackage[style=apa]{biblatex}
% \usepackage{natbib}
% \usepackage[font=footnotesize,labelfont=bf]{caption}
% \addbibresource{rmt.bib}
% \bibliographystyle{plain}
% \bibliography{rmt}
\begin{document}

@@

% \date{\today}

\keywords{random matrix theory, feature extraction, functional connectivity,
functional magnetic resonance imaging, voxelwise connectivity, predictive analysis}

\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse
    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make code publicly available to extract
    these features from a wide variety of data. We find preliminary evidence suggesting that
    RMT-inspired features may have unique potential in analyses of fMRI functional connectivity.
\end{abstract}

@@

In functional magnetic resonance imaging (fMRI), changes in the blood-oxygenation-level-dependent
(BOLD) signals are related to neural activity. It is common to investigate statistical relationships
between the BOLD signals through \textbf{functional connectivity} analyses, where correlations between
collections of these signals are examined to infer connections between different voxels or regions
of interest (ROIs) within the brain.

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
\citep{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}.
This complexity suggests fMRI is a candidate to be studied using \textbf{Random Matrix Theory} (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics \citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems
\citep{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\citep{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\citep{plerouRandomMatrixApproach2002} , social networks
\citep{jalanUncoveringRandomnessSuccess2014}, random networks
\citep{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in liquids
\citep{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}  and
amorphous clusters
\citep{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects of  amino acid
functional relationships \citep{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic
seizures \citep{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species \citep{agrawalQuantifyingRandomnessProtein2014} and breast
cancer \citep{raiRandomnessPreservedPatterns2015}.
\end{sloppypar}

@@

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states \citep{wangSpectralPropertiesTemporal2015}, between
subjects with and without ADHD \citep{wangRandomMatrixTheory2016}, and between pain and non-pain
states \citep{matharooSpontaneousBackpainAlters2020}. Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE. These findings suggest that
certain aspects of psychological processes might be characterized, in part, by features computed
from the eigenvalues of fMRI correlation matrices, and that these features might vary in an
interpretable manner across psychological processes. If this is the case, RMT could have unique
potential in characterizing the functioning of the human brain.

Here, we use a novel, voxel-based approach, and expand the applications of RMT to analyze fMRI scans
from  diverse datasets. We extract RMT-inspired whole-brain features from \textbf{voxelwise
functional connectivity} data, and assess the predictive value of these features. We also examine in
detail the robustness of the RMT methodology to various common analytic choices.

@@

As our primary intention is to examine the potential utility of RMT-derived features across a wide
variety of data, and not to decisively put forward or refute any specific hypotheses relating RMT to
psychological processes, we examine a wide selection of fMRI datasets available on the
\href{https://openneuro.org/}{OpenNeuro} platform [36].  This allows for a broad assessment of the
potential of voxelwise RMT features.

However, since previous studies employing RMT for feature extraction have partially interpreted
their findings with respect to attentional processes
\citep{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we have elected to use
datasets that involve attention in some way. This includes measures of attention or vigilance, or
comparisons between a condition involving more attention and focus (e.g. a task state) or conditions
which should involve considerably less attention or focus (e.g. resting-state).

@@

\emph{Trait Attention}.
The PANAS-X \citep{watsonPANASXManualPositive1994} is a 9-point Likert-type self-report measure
that queries the extent to which subjects agree particular affective terms have applied to them over
a recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) sum-score was constructed from
the scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects with a z-score greater / less than zero on this
sum-score were placed in the high / low ``trait attention'' category

\emph{Task Attention}.
The Conjunctive Continuous Performance Task (CCPT;
\cite{shalevConjunctiveContinuousPerformance2011} is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\emph{Vigilance}.
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \cite{gorgolewskiCorrespondenceIndividualDifferences2014}. One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in figures and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{table:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

@@

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast \citep{kunduMultiechoFMRIReview2017}. In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's \(i\)th signal value
from echo \(j\) as \(t_{j,i}\), then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in figures and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{table:1}. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
\citep[ANT;][]{fanActivationAttentionalNetworks2005}. Subgroup details are available in Table \ref{table:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in figures and tables with the label ``LEARN'') is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components \citep[see figure 1
of][]{schapiroHumanHippocampalReplay2018}. After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of
the analysis of this paper. Resting state scans were included in the low-attention ``rest''
subgroup.

@@

The RMT features extracted here are based on functional correlations rather than ROI activations,
and are completely insensitive to voxel location (spatial re-arrangement of an image’s voxels does
not change the eigenvalues of the correlation matrix). Thus RMT features summarize all voxel-wise
correlations in a whole-brain and location-free manner, which brings into question some of the usual
motivations for performing spatial smoothing, or registration to identify ROIs. In addition, if
image dimensions are identical between subjects, the RMT features (by virtue of their definition and
the \textbf{unfolding} procedure) already allow for direct comparisons. Thus, the usual motivations for
registration are also no longer clear when extracting RMT features.

@@

where \(\sigma^2\) is the true variance of the distribution of each i.i.d. entry (equal to 1 in the
case of correlation matrices), and where \(r = N / T\), for matrices of dimension \(N \times T\).
While real-world data will not, in general, be i.i.d., the extent to which spectra overlap with or
are well-described by the MP distribution has been used to guide analytical decisions regarding the
relevance of empirically observed eigenvalues. That is, since the majority of the eigenvalues of
random correlation matrices will fall within \([\lambda_{-}, \lambda_{+}]\), eigenvalues observed to
fall within this region are typically considered to be the result of noise
\citep{veraartDenoisingDiffusionMRI2016,veraartDiffusionMRINoise2016,ulfarssonDimensionEstimationNoisy2008,macmahonCommunityDetectionCorrelation2015}.
We examine as a potential feature the proportion of eigenvalues that fall within this range, and
call this the \textbf{noise ratio}. In addition, we calculate a broader, shifted noise ratio, based on the
adjustments of \cite{almogUncoveringFunctionalSignature2019} which takes into account the larger
degree of correlation in empirical data.

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table \ref{table:1}. The
various RMT features mentioned above were then extracted via the \texttt{empyricalRMT} library
\citep{bergerStfxecutablesEmpyricalRMTPreliminary2020} developed and made publicly available as part
of this study, as well as via code for this paper \citep{bergerderekm.DMBergerRandommatrixfmriV02020}.

@@

A wide variety of standard, out-of-the-box classifiers available in the Python scikit-learn library
\citep{pedregosaScikitlearnMachineLearning2011} were then employed to gauge the predictive value of
the RMT features. To allow for a wide range of flexibilities, classifiers employing k-nearest
neighbours (KNN), with k values in \(\{1, 2, 3, \dots, 10, 20\}\), logistic regression (LR), linear
discriminant analysis (LDA), and a support-vector machine (SVM) were employed. Default parameter
values were used when possible \citep[see][for precise function calls used]{bergerderekm.DMBergerRandommatrixfmriV02020}.

@@

\subsection{Overview}
Due to space limitations, we highlight only key results here. Extensive overviews of the full
results are available in the supplementary materials. Visual summaries of the prediction accuracies
for all features and analytic choices are presented in figures S1--4. Prediction accuracies for only the
multivariate RMT features are summarized in figures S5--7, and the distributions of the scalar features
are presented in figures S8--10. Prediction accuracies for the scalar features across analytic choices
are summarized in figures S11--12.

@@

With the exception of the LEARNING dataset, the RMT-inspired features all demonstrated some
predictive utility (figures S1-4). For example, there were strong differences in RMT features between
task and rest conditions for the REFLECT dataset, but the features were of variable predictive
utility in the PSYCH dataset, sometimes outperforming the naïve classifier, and sometimes not.
Although the RMT features were useful in the comparison involving duloxetine and controls (no pain)
in the OSTEO dataset, these features mostly did not aid in classifying between any of the other
possible pairings of subgroups in this dataset (figures S1-4).

Although the performance of classifiers trained on RMT features was variable, there was a tendency
for the raw and largest eigenvalues to be the most predictively useful features, especially in the
REFLECT and PARK data in discriminating between Parkinson's and controls (figures S5-7. S13; but see
Discussion).

\subsection{Raw Features}
As evident from figures S8-10, the scalar raw features (largest eigenvalue, noise ratio, and shifted
noise ratio) varied more within than between subgroups. They were not generally useful for
classifying conditions (Tables S15-17), with two or three notable exceptions.

The noise ratio (but not shifted noise ratio) differed consistently between the duloxetine and
control groups in the OSTEO data, allowing for prediction accuracies of up to 72\% compared to 51\%
for the naive classifier (figures S11-12 top-right panel, Table S5). The largest eigenvalue was also
useful in distinguishing between subjects self-reporting high versus low vigilance across sessions:
accuracies up to 64\% and 65\% (compared to naïve accuracies of 55\% and 50\%) for the first and
second sessions, respectively (Table S6). Finally, for the REFLECT data, task and rest were
perfectly distinguishable with various features and with some classifiers.

The sorted, raw eigenvalues, however, more consistently contained information that allowed some
separation of subgroups. Averaging across all datasets and comparisons, the raw eigenvalues
consistently demonstrated predictive utility (Tables S7--8). As visible in figures S13, the eigenvalues
had the most consistent predictive value for distinguishing individuals with Parkinson's from
controls, or distinguishing those with osteopathic pain taking duloxetine from controls (``nopain'' in
Fig. S13). The raw eigenvalues sometimes had predictive value for the PSYCH data, most notably in
comparing those differing in self-reported vigilance and high vs. low scorers in the attention task
(Table S7--8).

@@

\subsubsection{Role of Classifer}
The raw eigenvalues generally had the most predictive utility when employing either an SVM
classifier or KNN classifier with \(k > 3\) (Table S11, first row). When the raw eigenvalues were
first min-max normalized prior to prediction, an SVM classifier was clearly the most consistent in
predicting subgroup memberships (Table S12), however without normalization, KNN with \(k = 8\) or
\(k = 9\) generally best predicted subgroup membership across datasets (Table S11).

@@

The predictive utility of the unfolded RMT features varied considerably across datasets and
classifiers. However, some results were robust: for example, the spectral rigidity and level number
variance strongly differentiated controls from individuals prescribed duloxetine for osteopathic
pain, and this differentiation persisted regardless of unfolding degree or trimming
(\cref{fig:levelvar,fig:nnsd,fig:rigidity}, top-right panels).

@@

For other datasets, the predictive utility of the unfolded features was highly dependent on the
unfolding degree, and sometimes trim proportions. These differences are illustrated in detail for
the Parkinson's data in figures S14. For example, with minimal trimming, subgroup overlap in spectral
rigidity was minimized for unfolding degrees 5 and 9 (Fig. S14, first row), but with greater
trimming, this was not the case (Fig. S14, second row). For the level number variance with minimal
trimming, subgroup overlap appeared smaller for lower unfolding degrees (Fig. S14, third row),
however, with a greater degree of trimming, separation was clearer for slightly larger degrees (Fig.
S14, fourth row).

@@

\subsubsection{Role of Classifier}. Across all datasets, unfoldings, and trimmings, the spectral rigidity was
generally best able to separate between subgroups when either k was large (10 or 20) in KNN, or
Logistic Regression or SVM was employed (Table S13). The pattern was similar for the level number
variance, although the predictive utility of the level number variance was, overall, slightly more
variable than the spectral rigidity (Table S4).

@@

% \section{Conclusion}
